---
title: "Problem Set 1 — Problem 1"
format:
  html:
    code-fold: true
---

# Problem 1 — Exploratory Analysis of Soybean Cultivars

Place `data.csv` in the same folder as this `.qmd` file before rendering.

## Import and label the data

`data.csv` includes a header row. The names are checked and retained using the dataset documentation.

```{r}
#| message: false
#| warning: false

if (!file.exists("data.csv")) {
  stop("Place data.csv in the project folder before rendering this document.")
}

soy <- read.csv("data.csv", header = TRUE)
names(soy) <- c(
  "Season", "Cultivar", "Repetition", "PH", "IFP", "NLP",
  "NGP", "NGL", "NS", "MHG", "GY"
)

# Treat experimental labels as categorical variables.
soy$Season <- factor(soy$Season)
soy$Cultivar <- factor(soy$Cultivar)
soy$Repetition <- factor(soy$Repetition)

head(soy)
```

## Verify the data structure

```{r}
cat("Number of observations:", nrow(soy), "\n")
cat("Number of cultivars:", nlevels(soy$Cultivar), "\n")
cat("Number of seasons:", nlevels(soy$Season), "\n")

season_counts <- as.data.frame(table(Season = soy$Season))
names(season_counts)[2] <- "Observations"
season_counts
```

The checks above should show 320 observations, 40 cultivars, and two seasons.

## Correlation between plant height and grain yield

```{r}
overall_correlation <- cor(soy$PH, soy$GY)
overall_correlation

correlation_by_season <- aggregate(
  cbind(PH, GY) ~ Season,
  data = soy,
  FUN = mean
)

season_correlations <- aggregate(
  GY ~ Season,
  data = transform(soy, PH_GY_correlation = ave(
    PH, Season, FUN = function(x) NA_real_
  )),
  FUN = length
)

# A direct split-apply-combine calculation is clearer for correlations.
season_correlations <- data.frame(
  Season = levels(soy$Season),
  PH_GY_correlation = sapply(split(soy, soy$Season), function(d) cor(d$PH, d$GY))
)
season_correlations
```

The season with the larger `PH_GY_correlation` has the stronger positive association.

## Individual questions

```{r}
# Grain yield for the observation(s) with the largest plant height.
tallest_observation <- soy[soy$PH == max(soy$PH),
                           c("Season", "Cultivar", "Repetition", "PH", "GY")]
tallest_observation

# Percentage with MHG above the overall MHG average.
average_mhg <- mean(soy$MHG)
pct_mhg_above_average <- mean(soy$MHG > average_mhg) * 100
pct_mhg_above_average

# Cultivar(s) with the highest mean grain yield.
mean_gy_by_cultivar <- aggregate(GY ~ Cultivar, data = soy, FUN = mean)
mean_gy_by_cultivar <- mean_gy_by_cultivar[order(-mean_gy_by_cultivar$GY), ]
head(mean_gy_by_cultivar, 1)
```

## Mean of each quantitative variable

The variables below are agronomic measurements. Season, cultivar, and repetition are identifiers/design variables and are excluded.

```{r}
quantitative_vars <- c("PH", "IFP", "NLP", "NGP", "NGL", "NS", "MHG", "GY")

overall_means <- data.frame(
  Group = "Overall",
  t(colMeans(soy[quantitative_vars]))
)

season_means <- aggregate(
  soy[quantitative_vars],
  by = list(Season = soy$Season),
  FUN = mean
)
names(season_means)[1] <- "Group"
season_means$Group <- paste("Season", season_means$Group)

mean_table <- rbind(overall_means, season_means)
rownames(mean_table) <- NULL
mean_table
```

## Do grain yields differ by season?

Let $\mu_1$ and $\mu_2$ denote the population mean grain yields in Seasons 1 and 2.

- Null hypothesis: $H_0: \mu_1 = \mu_2$.
- Alternative hypothesis: $H_A: \mu_1 \ne \mu_2$.

```{r}
gy_t_test <- t.test(GY ~ Season, data = soy)
gy_t_test
```

Interpret the p-value using a 0.05 significance level: if the p-value is below 0.05, reject $H_0$ and conclude that mean grain yield differs between seasons. Otherwise, state that there is insufficient evidence of a difference in mean grain yield.
