---
title: "Problem Set 1 - Problem 3"
format:
  html:
    code-fold: true
    embed-resources: true
---

# Problem 3 - Happy Numbers

## Functions

```{r}
#' Calculate the next value in a happy-number sequence.
#'
#' @param n A single positive integer.
#' @return The sum of the squares of the digits of `n`.
nextHappy <- function(n) {
  if (!is.numeric(n) || length(n) != 1L || is.na(n) ||
      n <= 0 || n != floor(n)) {
    stop("n must be a single positive integer.")
  }

  digits <- as.integer(strsplit(as.character(n), "")[[1]])
  sum(digits ^ 2)
}

#' Return the happy-number sequence beginning at a given value.
#'
#' @param n A single positive integer.
#' @return A list with `sequence`, the visited values, and `isHappy`, which
#'   indicates whether the sequence reaches 1.
happySequence <- function(n) {
  if (!is.numeric(n) || length(n) != 1L || is.na(n) ||
      n <= 0 || n != floor(n)) {
    stop("n must be a single positive integer.")
  }

  sequence <- n
  seen <- integer(0)

  while (tail(sequence, 1) != 1 && !(tail(sequence, 1) %in% seen)) {
    seen <- c(seen, tail(sequence, 1))
    sequence <- c(sequence, nextHappy(tail(sequence, 1)))
  }

  list(sequence = sequence, isHappy = tail(sequence, 1) == 1)
}
```

## Required examples

```{r}
nextHappy(19)
nextHappy(82)
happySequence(19)
happySequence(4)
```

`nextHappy(19)` is 82 and `nextHappy(82)` is 68. The sequence for 19 is
19, 82, 68, 100, 1, so 19 is happy. The sequence starting at 4 cycles and
therefore is not happy.

## Happy numbers from 1 through 1000

```{r}
all_sequences <- lapply(1:1000, happySequence)
is_happy <- vapply(all_sequences, function(x) x$isHappy, logical(1))
sequence_lengths <- vapply(all_sequences, function(x) length(x$sequence), integer(1))

happy_count <- sum(is_happy)
happy_percent <- mean(is_happy) * 100
longest_length <- max(sequence_lengths)
longest_starts <- which(sequence_lengths == longest_length)

data.frame(
  happy_count = happy_count,
  happy_percent = happy_percent,
  longest_length = longest_length,
  longest_starting_values = paste(longest_starts, collapse = ", ")
)
```

There are 143 happy numbers between 1 and 1000, which is 14.3% of the
integers considered. The longest observed sequence has 20 displayed values
before reaching a repeated value; it occurs for starting values 269, 296,
629, 667, 676, 692, 766, 799, 926, 962, 979, and 997.
